---
title: "Untitled"
description: |
  A new article created using the Distill format.
author:
  - name: Nora Jones 
    url: https://example.com/norajones
    affiliation: Spacely Sprockets
    affiliation_url: https://example.com/spacelysprokets
date: "`r Sys.Date()`"
output: distill::distill_article
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

We want to better understanding changes in population and demographics
in small geographic areas of Madison, using Census data.

```{r}
library(tidyverse)
library(tidycensus)
library(tmap)
library(sf)
options(tigris_use_cache = TRUE)
```

Get data

# Identifying variables

For the 2020 Census data, variables are different, as the data released
is only in the format provided for redistricting. The `load_variables`
command from `tidycensus` loads all variables.

```{r}
variables_00 <- load_variables(2000, "sf1")
variables_10 <- tidycensus::load_variables(2010, "sf1")
variables_20 <- load_variables(2020, "pl")


```

Unfortunately, the variables do indeed change from each dataset to the
other.

For the 2020 data, the two relevant tables are P01 (Race) and P2
("Hispanic or Latino, and Not Hispanic or Latino by Race") for the total
population. `P1_001N` is the total population. `P1_003N` is
`White alone`. `P1_004N` is Black/African American alone. `P1_006N`
Asian alone. `P2_005N` is non-Hispanic/Latino White. `P2_002N` is
Hispanic/Latino of any race.

For 2010:

-   P001001: Total population (to be used as the `summary_var` for
    percentages)
-   P005002: Not Hispanic/Latino, of any race
-   P005003: Non-Hispanic/Latino White
-   P005010: Hispanic/Latino of any race (same as P004003)
-   P003002: White alone
-   P003003: Black alone
-   P003005: Asian alone

And for 2000: - P001001: Total population (to be used as the
`summary_var` for percentages) - P004003: Not Hispanic/Latino, of any
race - P004005: Non-Hispanic/Latino White - P004002: Hispanic/Latino of
any race - P003003: White alone - P003004: Black alone - P003006: Asian
alone

```{r}
vars20 <- c("P1_001N", "P1_003N", "P1_004N", "P1_006N", "P2_005N", "P2_002N")
vars10 <- c("P001001", "P005003", "P005010", "P003002", "P003003", "P003005")
vars00 <- c("P001001", "P004003", "P004005", "P004002", "P003003", "P003004", "P003006")
```

Have to make decisions on block versus block group. If going with block,
need to filter data to make manageable. Block groups seem a little
coarse for certain purposes.

```{r}
dane_2020 <- get_decennial(
  geography = "block group",
  variables = vars20, 
  year = 2020,
  state = "WI",
  county = "Dane",
  geometry = T,
  summary_var = "P1_001N") %>%
  filter(summary_value != 0) %>%
  mutate(pct = 100 * (value / summary_value), 
         variable = case_when(
           variable == "P1_001N" ~ "Total",
           variable == "P1_003N" ~ "White alone",
           variable == "P1_004N" ~ "Black/African American alone", 
           variable == "P1_006N" ~ "Asian alone", 
           variable == "P2_005N" ~ "Non-Hispanic White",
           variable == "P2_002N" ~ "Hispanic of any race"),
         year = 2020
         )


dane_2010 <- get_decennial(
  geography = "block group",
  variables = vars10, 
  year = 2010,
  state = "WI",
  county = "Dane",
  geometry = T,
  summary_var = "P001001") %>%
  filter(summary_value != 0) %>%
  mutate(pct = 100 * (value / summary_value),
         variable = case_when(
           variable == "P001001" ~ "Total",
           variable == "P003002" ~ "White alone",
           variable == "P003003" ~ "Black/African American alone",
           variable == "P003005" ~ "Asian alone", 
           variable == "P005003" ~ "Non-Hispanic White",
           variable == "P005010" ~ "Hispanic of any race"),
         year = 2010
         )

dane_2000 <- get_decennial(
  geography = "block group",
  variables = vars00, 
  year = 2000,
  state = "WI",
  county = "Dane",
  geometry = T,
  summary_var = "P001001") %>%
  filter(summary_value != 0) %>%
  mutate(pct = 100 * (value / summary_value),
         variable = case_when(
           variable == "P001001" ~ "Total", 
           variable == "P003003" ~ "White alone",
           variable == "P003004" ~ "Black/African American alone",
           variable == "P003006" ~ "Asian alone", 
           variable == "P004005" ~ "Non-Hispanic White",
           variable == "P004002" ~ "Hispanic of any race"),
         year = 2000
         )


# 
# tmap_mode("view")
# dane_2020 %>%
#   tm_shape() +
#   tm_polygons("pct") +
#   tm_facets("variable")
  
```

Now we combine the three datasets:

```{r}
dane_bg <- bind_rows(dane_2000, dane_2010, dane_2020)
```

Faceted plot of Black alone:

```{r}
tmap_mode("view")
dane_bg %>% 
  filter(variable == "Hispanic of any race") %>% 
  tm_shape() +
  tm_polygons("pct", alpha = .3) +
  tm_facets("year", sync = T)
```

# Dealing with changing geographies

Comparing the number of observations between the three years shows that
the number of block groups has grown:

```{r}
dane_bg %>% 
  st_drop_geometry() %>% 
  group_by(year, variable) %>% 
  summarize(n())
```

In the 2000 census there were only 258 block groups, whereas in 2020
there were almost 100 more. This reflects the growth in population in
Dane County.

```{r}

```

Let's look at relationship files directly from the Census Bureau:

```{r}
relationships <- read_delim("data/tab2010_tab2020_st55_wi.txt", "|")
```

To aggregate at 2010 levels, we would:

1.  Create GEOIDs for 2010 and 2020 in the relationship dataframe
2.  Join `dane_2020 with`relationship`by`GEOID_20\`
3.  Group by `GEOID_10`
4.  Summarize counts

As a check, we can see how many GEOIDs remain and if they match the
number in `dane_2010`.

```{r}
x <- relationships %>% 
  filter(COUNTY_2010 == "025") %>% 
  mutate(geoid_2010 = paste0(STATE_2010, COUNTY_2010, TRACT_2010, BLK_2010),
         geoid_2020 = paste0(STATE_2020, COUNTY_2020, TRACT_2020, BLK_2020)) %>% 
    filter(is.na(BLOCK_PART_FLAG_O) & is.na(BLOCK_PART_FLAG_R)) %>% 
  right_join(dane_2020, by = c("geoid_2020" = "GEOID"))
  tm_shape()+
  tm_polygons("red")
  filter(geoid_2010 != geoid_2020) %>% 
  view()


relationships %>% 
  filter(COUNTY_2010 == "025") %>% 
  filter(is.na(BLOCK_PART_FLAG_O) & is.na(BLOCK_PART_FLAG_R)) %>% 
  left_join(dane_2020, by = c("geoid_2020" = "GEOID"))
```

Unfortunately, it's even more complicated than this. As the
documentation puts it: "Block relationships may be one-to-one,
many-to-one, one-to-many, or many-to-many."[@bureau]

An easy way to deal with this is use NHGIS Geographic Crosswalks. To
allow for comparisons over time, the crosswalks provides interpolation
weights that "indicates the proportion of a source zone's
characteristics that should be allocated to a specific target
zone."[@schroeder2007]

To get a better sense of what these changes look like geographically, I
will explore the geographies in some more detail.

```{r}
blocks2000 <- tigris::blocks(55, county = "025", year = 2000) %>% 
  mutate(year = 2000) %>% 
  rename(GEOID = BLKIDFP00,
         ALAND = ALAND00) %>% 
  select(GEOID, ALAND, year, geometry)

blocks2010 <- tigris::blocks(55, county = "025", year = 2010) %>% 
  mutate(year = 2010) %>% 
  rename(GEOID = GEOID10,
         ALAND = ALAND10) %>% 
  select(GEOID, ALAND, year, geometry)


blocks2020 <- tigris::blocks(55, county = "025", year = 2020) %>% 
  mutate(year = 2020) %>% 
  rename(GEOID = GEOID10,
         ALAND = ALAND10) %>% 
  select(GEOID, ALAND, year, geometry)

blocks <- rbind(blocks2000, blocks2010, blocks2020)

```

```{r}
tmap_mode("view")
tm_shape(blocks) +
  tm_polygons() +
  tm_facets("year", sync = T)
```
